Testing for measurement invariance with respect to an ordinal variable

### 2012-08-10 ordinal/ordinal.Rnw
Additions to manuscript: o Added application section using youth gratitude data. o Expanded outline for general discussion. o General revisions to paper.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@158 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

%% TODO add some more citations to these initial paragraphs
The study of measurement invariance and differential item functioning
(DIF) 
has received considerable attention in the psychometric literature
(see, e.g., \citeNP{Mill11} for a thorough review).  A set of
psychometric scales ${\bm X}$ are defined to be measurement invariant
with respect to an auxiliary variable $V$ if \cite{Mel89}
\begin{equation}
    \label{eq:midef}
      f(x_i | t_i, v_i, \dots) = f({\bf x}_i | t_i, \dots),
\end{equation}
where $T$ is the latent variable that the scales measure, $f$ is the
model's distributional form, and the $i$ subscript refers to
individual cases.  If the above equation does not hold, then there
exists a measurement invariance violation.  We focus here on
situations where $f()$ is parametric, and the measurement invariance
violation occurs because the model parameters are
unequal across individuals (and related to $V$).

As a concrete example of the study of measurement invariance, consider
a situation where ${\bm X}$ contains some ``high stakes'' tests of
ability and $V$ is ethnicity.
One's ethnicity should be unrelated to the measurement parameters
within $f()$, and this expectation can be studied within a
psychometric model by fitting the model examining whether or not
measurement parameters 
vary across different ethnicities.  Statistical tools that can be used
to carry out this study include Likelihood Ratio Tests, Lagrange
Multiplier tests, and Wald tests \cite{Sat89}.
These tools have greatly aided in
the development of improved, ``fairer'' psychometric tests and
scales.

@@

@@

In studying measurement invariance, we are essentially testing the
assumption that all  
individuals arise from the same 
parameter vector ${\bm \theta}$.
Thus, a hypothesis of invariance can be written as
\begin{equation}
    \label{eq:h0}
    H_0:~ {\bm \theta}_i = {\bm \theta}_0,\quad (i=1,\ldots,n),
\end{equation}
where ${\bm \theta}_i$ reflects the parameter vector for individual
$i$ (and modifications for subsets of ${\bm \theta}$ are immediate).  This
hypothesis is typically tested against an alternative
\begin{equation}
  \label{eq:h1*}
  H_1^*:~ {\bm \theta}_i = \left\{ \begin{array}{ll}
    {\bm \theta}^{(A)} & \mbox{if } v_i \le \nu, \\
    {\bm \theta}^{(B)} & \mbox{if } v_i >   \nu,
  \end{array} \right.
\end{equation}
where $v_i$ reflects individual $i$'s value on an auxiliary variable
$V$ and $\nu$ is a threshold dividing individuals into groups based on
$V$.  In this paper, we generally focus on situations where $V$ is
ordinal and 
where $\nu$ is unknown.  In the section below, we review tests where
$V$ is continuous (and $\nu$ is unknown) before proceeding to the 
proposed tests for ordinal $V$.

@@

When $V$ is categorical with a relatively-small number of categories,
tests of measurement invariance typically proceed via multiple-group
models.  In this situation, we use likelihood ratio tests to compare a
model whose parameters differ across groups to a model whose
parameters are constrained to be equal across groups.  When $V$ is
continuous, however, multiple-group models obviously cannot be used 
because there are no existing groups.  Instead, we can fit a model
whose parameters are restricted to be equal across all individuals, then
examine individuals' scores ${\bm s}(\hat{{\bm \theta}}; x_i)$
fluctuate with their values of $V$.  If measurement invariance holds
with respect to $V$, then the scores should randomly fluctuate around
zero.  Conversely, if measurement invariance does not hold, then the
scores should systematically depart from zero.  These ideas are
related to those underlying the Lagrange multiplier test; see
\citeNP{MerZei12} for more details.

To formalize these ideas, we define the $k$-dimensional {\em
  cumulative score process} as 
\begin{equation} \label{eq:cumscore}
  {\bm B}(t; \hat {\bm \theta}) ~=~ \hat {\bm I}^{-1/2} n^{-1/2}
    \sum_{i = 1}^{\lfloor n t \rfloor} {\bm s}(\hat {\bm \theta}; x_i)
  \qquad (0 \le t \le 1)
\end{equation}
where $\lfloor nt \rfloor$ is the integer part of $nt$ and
$\hat I$ is some consistent estimate of the covariance matrix of the
scores.  Equation~\eqref{eq:cumscore} simultaneously accounts for the
ordering of individuals w.r.t.\ $V$ and decorrelates the scores
associated with each of the $k$ model parameters (which allows us to
potentially make inferences separately for each individual model
parameter).  Using ideas similar to those that were outlined in the previous
paragraph, the cumulative score process associated with each
model parameter should randomly fluctuate around zero under
measurement invariance.  Further, there exists a functional central
limit theorem that allows us to make formal inference with this
cumulative score process.  Assuming that individuals are independent
and the usual ML regularity conditions hold, it is possible to show
that \citeA{HjoKon02}
\begin{equation} \label{eq:fclt}
  {\bm B}(\cdot; \hat {\bm \theta}) ~\overset{d}{\rightarrow}~ {\bm B}^{0}(\cdot),
\end{equation}
where $\overset{d}{\rightarrow}$ denotes convergence in distribution
and ${\bm B}^{0}(\cdot)$ is a $k$-dimensional Brownian bridge.
Thus, we can construct tests of measurement invariance by comparing
the cumulative score process's behavior to that of a Brownian bridge.
This is accomplished by comparing a scalar statistic associated with
the cumulative score process to the analogous statistic of a Brownian
bridge.

@@

Test statistics with higher power
aggregate information across parameters and possibly also across
individuals.  These test statistics include
\begin{eqnarray}
    \label{eq:cvm}
    \mathit{CvM}     & = & n^{-1} \sum_{i = 1,\dots, n} \sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2, \\
    \label{eq:maxlm}
    \max \mathit{LM} & = & \max_{i = \underline{i}, \dots, \overline{\imath}} ~
      \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) \right\}^{-1}
      \sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2,
\end{eqnarray}
with the former being a Cram{\'e}r-von Mises statistic and the latter
corresponding to a ``maximum'' Lagrange multiplier test, where the
maximum is taken across all possible divisions of individuals into two
groups w.r.t.\ $V$.  Additionally, the $\max \mathit{LM}$ statistic is
scaled by the asymptotic variance $t (1 - t)$ of the process ${\bm
  B}(t, \hat {\bm \theta})$.  In simulations, Merkle and Zeileis found that the
$\mathit{CvM}$ test exhibited the highest power for testing multiple
factor analysis parameters for measurement invariance w.r.t.\
continuous $V$.  These simulations included situations in which
subsets of model parameters were tested; such situations are handled
by focusing only on those columns of ${\bm B}(\hat {\bm \theta})_{ij}$
that correspond to the parameters of interest.

@@

The theory described above was designed for situations where
$V$ is continuous, so that there is a unique
ordering of individuals with respect to $V$.  In
situations where $V$ is categorical, it is possible
to obtain a test statistic by first summing scores within each of the
$m$ levels of the
auxiliary variable, then ``summing the sums'' to obtain a test
statistic \cite{HjoKon02}.  This test statistic can be formally
written as
\begin{equation}
    \label{eq:lmuo}
     \mathit{LM}_\mathit{uo} = \sum_{\ell = 1, \dots, m} \sum_{j = 1, \dots, k}
       \left( {\bm B}(\hat {\bm \theta})_{i_\ell j} - {\bm B}(\hat {\bm \theta})_{i_{\ell - 1}j} \right)^2,
\end{equation}
where, again, tests of subsets of model parameters can be obtained by
taking the inner sum over only the $k^* < k$ parameters of interest.
This test statistic discards the ordinal nature of the auxiliary
variable.  A similar issue is observed in testing for measurement
invariance via multiple groups models and Likelihood Ratio tests (or,
equivalently, via Wald tests or Lagrange multiplier tests): we
can allow $\bm{\theta}$ to be unique at each level of the ordinal
variable, but the ordinality of the auxiliary variable is lost.
In contrast, the statistics proposed below explicitly account for the fact
that $V$ is ordinal.

The statistics proposed here are
similar to those described in
Equations~\eqref{eq:dmax},~\eqref{eq:cvm}, and~\eqref{eq:maxlm}
above, except that we focus on ``bins'' of individuals at each level of the
ordinal variable.  That is, instead of aggregating over all $i = 1,
\dots, n$ individuals, we first compute cumulative proportions
$t_\ell$ ($\ell = 1, \dots, m-1$) associated with the first $m-1$
levels of $V$.  We then aggregate the 
cumulative scores only over $i_\ell = \lfloor n
\cdot t_\ell\rfloor$.  Test statistics related to~\eqref{eq:dmax}
and~\eqref{eq:maxlm} above can then be written as
%% TODO insert equation
\begin{eqnarray}
    \label{eq:wdm}
      \mathit{WDM}_o & = & \max_{i \in \{i_1, \dots, i_m \}} ~ \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) \right\}^{-1}      
                             \max_{j = 1, \dots, k} | {\bm B}(\hat {\bm \theta})_{ij} |,\\
    \label{eq:maxlmo}
      \max \mathit{LM}_o & = & \max_{i \in \{i_1, \dots, i_m \}} ~
                             \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) \right\}^{-1}
                             \sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2.\\
\end{eqnarray}
Critical values associated with these test statistics can be obtained
by applying the same functionals to bins of a Brownian bridge, where the
bin sizes result in the cumulative proportions $t_\ell$ ($\ell=1, \dots,
m-1$) associated with the observed $V$.
Closed-form solutions for
these particular functionals of a Brownian bridge are unavailable, but
critical values and $p$-values can be obtained through repeated
simulation of a Brownian bridge.  This functionality is built in to
R's strucchange package \cite{strucchange}, which can be used to
generally carry out the tests.

As we demonstrate in the simulations below, the test
statistics proposed above are sensitive to the measurement invariance
violations that 
an analyst would typically expect from an ordinal $V$.
These include violations that increase or decrease as we move up
levels of $V$, as well as violations that abruptly
begin at a single point on $V$ and continue after
that.  The test statistics are insensitive to anomolies in a small
number of categories of $V$ that are
unrelated to the ordering of $V$.  This is especially relevant to situations
where the analyst has a large sample size, where the usual
Likelihood Ratio Test is notoriously sensitive to minute parameter
instabilities (e.g., \citeNP{Ben80}).

@@

The more interesting result of this simulation lies in the comparison
of the four test statistics within each panel of the figure.  The
two ``unordered'' test statistics both have relatively-high power to
detect the measurement invariance violation, illustrating the finding
of Bentler and Bonett \citeyear{BenBon80} and others that the
likelihood ratio test statistic picks out minor 
parameter discrepancies at large $n$.  In contrast, the two ordinal
test statistics that we proposed have considerably lower power, with
the $\text{WDM}_O$ statistic being the lowest and the $\text{LM}_O$
statistic being higher at larger values of $d$.  
This demonstrates that the proposed ordinal test
statistics can be especially useful at large sample sizes, where
traditional test statistics result in frequent significance.
Both statistics exhibited much lower power to detect a measurement
invariance violation occurring at a single level of the auxiliary
variable.

@@

@@

Froh et al.\ \citeyear{FroFan11} used confirmatory factor analysis
models to study the invariance of three youth gratitude scales across
students aged 10 to 19 years.  Each of the scales were individually
factor-analyzed using the items comprising the scale.
For each model, the authors 
first fit a congeneric model (all parameters free for each age group),
followed by a tau-equivalent model (factor loadings restricted to be
equal across groups) and a parallel model (all parameters restricted
to be equal across groups).  Because their sample size was large ($n
\approx 1400$), they could not rely solely on likelihood ratio tests
(i.e., chi-square difference tests) for model comparison because the
tests were overly sensitive at their sample size.  To supplement these
tests, Froh et al examined a set of alternative fit indices, including
the non-normed fit index, the comparative fit index, and the
incremental fit index. %% TODO cite
The authors generally found support for the tau-equivalent models
through these alternative fit indices; the likelihood ratio test often
resulted in significance event when the alternative indices indicated
good fit.

In the analyses described below, we re-analyze the Froh et al.\ data using
the ordinal test statistics proposed in this paper.  This results in a
series of tests that are less sensitive to sample size than the
likelihood ratio test, while being
more sensitive to ``ordinal'' violations of measurement invariance.
We focus on two analyses from Froh et al.\ where the likelihood
ratio test resulted in significance (indicating that the restricted
model did not fit as well as the less-restricted model), but the
alternative fit measures indicated the opposite (using the rules of
thumb employed by Froh et al).  These include
comparison of a one-factor congeneric model to a one-factor
tau-equivalent model using the 
GQ6 and comparison of a one-factor tau-equivalent model to a one-factor
parallel model 
using the GAC.  To conduct equivalent analyses via the proposed tests, we
fit the restricted model and test for instability in the focal
model parameters.

@@

The results section is divided into two subsections, one for each of
the analyses described previously.

@@

\paragraph{GQ6.} Ordinal tests do not support LRT.

@@

\paragraph{GAC.} Ordinal tests do support LRT.

@@

\begin{enumerate}
  \item The only proposed statistics that treat ordinal auxiliary
    variables as truly ordinal?  LRT treats them as categorical,
    inclusion in the model typically treats tham as continuous.
  \item The first point is important because, if we test wrt an
    ordinal auxiliary variable, we are generally looking for ordinal
    violations.  We don't care if a middle level differs from
    all others because we would not usually have an intuitive
    explanation for it.
  \item The proposed statistics should generally transfer to IRT and
    other psychometric models.
\end{enumerate}

>>>>

%% TODO add some more citations to these initial paragraphs
The study of measurement invariance and differential item functioning
(DIF) 
has received considerable attention in the psychometric literature
(see, e.g., \citeNP{Mil11} for a thorough review).  A set of
psychometric scales ${\bm X}$ are defined to be measurement invariant
with respect to an auxiliary variable $V$ if \cite{Mel89}
\begin{equation}
    \label{eq:midef}
      f(x_i | t_i, v_i, \dots) = f({\bf x}_i | t_i, \dots),
\end{equation}
where $T$ is the latent variable that the scales measure, $f$ is the
model's distributional form, and the $i$ subscript refers to
individual cases.  If the above equation does not hold, then there
exists a measurement invariance violation.  We focus here on
situations where $f()$ is parametric, and the measurement invariance
violation occurs because the model parameters are
unequal across individuals (and related to $V$).

As a concrete example of the study of measurement invariance, consider
a situation where ${\bm X}$ contains some ``high stakes'' tests of
ability and $V$ is ethnicity.
One's ethnicity should be unrelated to the measurement parameters
within $f()$, and this expectation can be studied within a
psychometric model by fitting the model and examining whether or not
measurement parameters 
vary across different ethnicities.  Statistical tools that can be used
to carry out this study include Likelihood Ratio Tests, Lagrange
Multiplier tests, and Wald tests (e.g., \citeNP{Sat89}).
These tools have greatly aided in
the development of improved, ``fairer'' psychometric tests and
scales.

@@

This section contains background on the theory underlying the
proposed statistics; for a more detailed
account, see Merkle and Zeileis \citeyear{MerZei12}.

@@

In studying measurement invariance, we are essentially testing the
assumption that all  
individuals arise from the same 
parameter vector ${\bm \theta}$.
Thus, a hypothesis of invariance can be written as
\begin{equation}
    \label{eq:h0}
    H_0:~ {\bm \theta}_i = {\bm \theta}_0,\quad (i=1,\ldots,n),
\end{equation}
where ${\bm \theta}_i$ reflects the parameter vector for individual
$i$ (and modifications for subsets of ${\bm \theta}$ are immediate).  This
hypothesis is often tested against an alternative
\begin{equation}
  \label{eq:h1*}
  H_1^*:~ {\bm \theta}_i = \left\{ \begin{array}{ll}
    {\bm \theta}^{(A)} & \mbox{if } v_i \le \nu, \\
    {\bm \theta}^{(B)} & \mbox{if } v_i >   \nu,
  \end{array} \right.
\end{equation}
where $v_i$ reflects individual $i$'s value on an auxiliary variable
$V$ and $\nu$ is a threshold dividing individuals into groups based on
$V$.  Note that this alternative hypothesis can also be used when $V$
is ordinal, testing whether the model parameters shift at some point
on $V$.  
Alternatively, if $V$ is categorical with a relatively-small number of
groups, a frequent alternative of interest is
\begin{equation}
  \label{eq:h2*}
  H_2^*:~ {\bm \theta}_i = {\bm \theta}_{v_i}.
\end{equation}
In this paper, we generally focus on situations where $V$ is
ordinal and where the measurement invariance violation is related to the ordinal
variable (e.g., the violation is of the type from~\eqref{eq:h1*} or
the violation grows/shrinks with $V$).  Researchers typically test for
measurement invariance w.r.t.\ ordinal $V$ by employing the alternative
from~\eqref{eq:h2*}, which implicitly treats $V$ as categorical.
Thus, test statistics that explicitly treat $V$ as ordinal should have
higher power to detect the ``ordinal'' types of violations described above.

In the section below, we review tests where
$V$ is continuous (and $\nu$ is unknown) before proceeding to the 
proposed tests for ordinal $V$.

@@

When $V$ is categorical with a relatively-small number of categories,
tests of measurement invariance typically proceed via multiple-group
models.  In this situation, we use likelihood ratio tests to compare a
model whose parameters differ across groups to a model whose
parameters are constrained to be equal across groups.  When $V$ is
continuous, however, multiple-group models obviously cannot be used 
because there are no existing groups.  Instead, we can fit a model
whose parameters are restricted to be equal across all individuals, then
examine how individuals' scores ${\bm s}(\hat{{\bm \theta}}; x_i)$
fluctuate with their values of $V$.  If measurement invariance holds
with respect to $V$, then the scores should randomly fluctuate around
zero.  Conversely, if measurement invariance does not hold, then the
scores should systematically depart from zero.  These ideas are
related to those underlying the Lagrange multiplier test; see
\citeNP{MerZei12} for more details.

To formalize these ideas, we define the $k$-dimensional {\em
  cumulative score process} as 
\begin{equation} \label{eq:cumscore}
  {\bm B}(t; \hat {\bm \theta}) ~=~ \hat {\bm I}^{-1/2} n^{-1/2}
    \sum_{i = 1}^{\lfloor n t \rfloor} {\bm s}(\hat {\bm \theta}; x_i)
  \qquad (0 \le t \le 1)
\end{equation}
where $\lfloor nt \rfloor$ is the integer part of $nt$ and
$\hat I$ is some consistent estimate of the covariance matrix of the
scores.  Equation~\eqref{eq:cumscore} simultaneously accounts for the
ordering of individuals w.r.t.\ $V$ and decorrelates the scores
associated with each of the $k$ model parameters (which allows us to
potentially make inferences separately for each individual model
parameter).  Using ideas similar to those that were outlined in the previous
paragraph, the cumulative score process associated with each
model parameter should randomly fluctuate around zero under
measurement invariance.  Further, there exists a functional central
limit theorem that allows us to make formal inference with this
cumulative score process.  Assuming that individuals are independent
and the usual ML regularity conditions hold, it is possible to show
that \cite{HjoKon02}
\begin{equation} \label{eq:fclt}
  {\bm B}(\cdot; \hat {\bm \theta}) ~\overset{d}{\rightarrow}~ {\bm B}^{0}(\cdot),
\end{equation}
where $\overset{d}{\rightarrow}$ denotes convergence in distribution
and ${\bm B}^{0}(\cdot)$ is a $k$-dimensional Brownian bridge.
Thus, we can construct tests of measurement invariance by comparing
the cumulative score process's behavior to that of a Brownian bridge.
This is accomplished by comparing a scalar statistic associated with
the cumulative score process to the analogous statistic of a Brownian
bridge.

@@

Test statistics that exhibit better performance in such situations 
aggregate information across parameters and possibly also across
individuals.  These test statistics include
\begin{eqnarray}
    \label{eq:cvm}
    \mathit{CvM}     & = & n^{-1} \sum_{i = 1,\dots, n} \sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2, \\
    \label{eq:maxlm}
    \max \mathit{LM} & = & \max_{i = \underline{i}, \dots, \overline{\imath}} ~
      \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) \right\}^{-1}
      \sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2,
\end{eqnarray}
with the former being a Cram{\'e}r-von Mises statistic and the latter
corresponding to a ``maximum'' Lagrange multiplier test, where the
maximum is taken across all possible divisions of individuals into two
groups w.r.t.\ $V$.  Additionally, the $\max \mathit{LM}$ statistic is
scaled by the asymptotic variance $t (1 - t)$ of the process ${\bm
  B}(t, \hat {\bm \theta})$.  In simulations, Merkle and Zeileis found that the
$\mathit{CvM}$ test exhibited the highest power for testing multiple
factor analysis parameters for measurement invariance w.r.t.\
continuous $V$.  These simulations included situations in which
subsets of model parameters were tested; such situations are handled
by focusing only on those columns of ${\bm B}(\hat {\bm \theta})_{ij}$
that correspond to the parameters of interest.

@@

The theory described above was designed for situations where
$V$ is continuous, so that there is a unique
ordering of individuals with respect to $V$.  In
situations where $V$ is categorical, it is possible
to obtain a test statistic by first summing scores within each of the
$m$ levels of the
auxiliary variable, then ``summing the sums'' to obtain a test
statistic \cite{HjoKon02}.  This test statistic can be formally
written as
\begin{equation}
    \label{eq:lmuo}
     \mathit{LM}_\mathit{uo} = \sum_{\ell = 1, \dots, m} \sum_{j = 1, \dots, k}
       \left( {\bm B}(\hat {\bm \theta})_{i_\ell j} - {\bm B}(\hat {\bm \theta})_{i_{\ell - 1}j} \right)^2,
\end{equation}
where, again, tests of subsets of model parameters can be obtained by
taking the inner sum over only the $k^* < k$ parameters of interest.
This test statistic discards the ordinal nature of the auxiliary
variable, essentially employing the alternative hypothesis
from~\eqref{eq:h2*}.  A similar issue is observed in testing for
measurement 
invariance via multiple groups models and Likelihood Ratio tests (or,
equivalently, via Wald tests or Lagrange multiplier tests): we
can allow $\bm{\theta}$ to be unique at each level of the ordinal
variable, but the ordinality of the auxiliary variable is lost.
In contrast, the statistics proposed below explicitly account for the fact
that $V$ is ordinal.

The statistics proposed here are
similar to those described in
Equations~\eqref{eq:dmax} and~\eqref{eq:maxlm}
above, except that we focus on ``bins'' of individuals at each level of the
ordinal variable.  That is, instead of aggregating over all $i = 1,
\dots, n$ individuals, we first compute cumulative proportions
$t_\ell$ ($\ell = 1, \dots, m-1$) associated with the first $m-1$
levels of $V$.  We then aggregate the 
cumulative scores only over $i_\ell = \lfloor n
\cdot t_\ell\rfloor$.  Test statistics related to~\eqref{eq:dmax}
and~\eqref{eq:maxlm} above can then be written as
%% TODO insert equation
\begin{eqnarray}
    \label{eq:wdm}
      \mathit{WDM}_o & = & \max_{i \in \{i_1, \dots, i_m \}} ~ \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) \right\}^{-1}      
                             \max_{j = 1, \dots, k} | {\bm B}(\hat {\bm \theta})_{ij} |,\\
    \label{eq:maxlmo}
      \max \mathit{LM}_o & = & \max_{i \in \{i_1, \dots, i_m \}} ~
                             \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) \right\}^{-1}
                             \sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2.
\end{eqnarray}
Critical values associated with these test statistics can be obtained
by applying the same functionals to bins of a Brownian bridge, where the
bin sizes result in the cumulative proportions $t_\ell$ ($\ell=1, \dots,
m-1$) associated with the observed $V$.
Closed-form solutions for
these particular functionals of a Brownian bridge are unavailable, but
critical values and $p$-values can be obtained through repeated
simulation of a Brownian bridge.  This functionality is built in to
R's strucchange package \cite{ZeiLei02}, which can be used to
generally carry out the tests.

As demonstrated in the simulations below, the test
statistics proposed above are sensitive to the measurement invariance
violations that 
an analyst would typically expect from an ordinal $V$.
These include violations that increase or decrease as we move up
levels of $V$, as well as violations that abruptly
begin at a single point on $V$ and continue after
that.  The test statistics are insensitive to anomolies in a small
number of categories of $V$ that are
unrelated to the ordering of $V$.  This is especially relevant to situations
where the analyst has a large sample size, where the usual
Likelihood Ratio Test is notoriously sensitive to minute parameter
instabilities (e.g., \citeNP{Ben80}).

@@

The more interesting result of this simulation lies in the comparison
of the four test statistics within each panel of the figure.  The
two ``unordered'' test statistics both have relatively-high power to
detect the measurement invariance violation, illustrating the finding
of Bentler and Bonett \citeyear{Ben80} and others that the
likelihood ratio test statistic picks out minor 
parameter discrepancies at large $n$.  In contrast, the two ordinal
test statistics that we proposed have considerably lower power, with
the $\text{WDM}_O$ statistic being the lowest and the $\text{LM}_O$
statistic being higher at larger values of $d$.  
This demonstrates that the proposed ordinal test
statistics can be especially useful at large sample sizes, where
traditional test statistics result in frequent significance.
Both statistics exhibited much lower power to detect a measurement
invariance violation occurring at a single level of the auxiliary
variable.

@@

\readme{Fan: Can you provide a short
  background on the study of youth gratitude and the specific intent
  of Froh et al?}

@@

Froh et al.\ \citeyear{FroFan11} used confirmatory factor analysis
models to study the invariance of three youth gratitude scales across
students aged 10 to 19 years.  Due to sample size constraints, the age
variable included six categories: 10-11 years, 12-13 years, 14 years,
15 years, 16 years, and 17-19 years.  Thus, age is an ordinal variable
with which the proposed tests can be applied.

To test for measurement invariance w.r.t.\ age, each of the three scales
was individually 
factor-analyzed using the items that comprised the scale.
For each model, the authors 
first fit a congeneric model (all parameters free for each level of age),
followed by a tau-equivalent model (factor loadings restricted to be
equal across each level of age) and a parallel model (all parameters restricted
to be equal across levels of age).  Because their sample size was large ($n
\approx 1400$), they could not rely solely on likelihood ratio tests
(i.e., chi-square difference tests) for model comparison because the
tests were overly sensitive at their sample size.  To supplement these
tests, Froh et al examined a set of alternative fit indices, including
the non-normed fit index, the comparative fit index, and the
incremental fit index. %% TODO cite
The authors generally found support for the tau-equivalent models
through these alternative fit indices; the likelihood ratio test often
resulted in significance event when the alternative indices indicated
good fit.

In the analyses described below, we re-analyze the Froh et al.\ data using
the ordinal test statistics proposed in this paper.  This results in a
series of tests that are less sensitive than the likelihood ratio test
to minor parameter discrepancies,
while being more sensitive to ordinal violations of measurement invariance.
We focus on two analyses from Froh et al.\ where the likelihood
ratio test resulted in significance (indicating that the restricted
model did not fit as well as the less-restricted model), but the
alternative fit measures employed by Froh et al.\ indicated the opposite.
These include
comparison of a one-factor congeneric model to a one-factor
tau-equivalent model using the 
GQ6 and comparison of a one-factor tau-equivalent model to a one-factor
parallel model 
using the GAC.  To conduct equivalent analyses via the proposed tests, we
fit the more-restricted model in each case and test for instability in
the focal model parameters.

@@

The results section is divided into two subsections, one for each of
the analyses described previously.  The first section contains an
example of the ordinal statistics disagreeing with the likelihood
ratio tests, while the second section contains the opposite.

@@

\subsubsection{GQ-6} In fitting a tau-equivalent model and a congeneric
model to the GQ-6 data, Froh et al.\ used alternative fit indices to
conclude that the tau-equivalent model was as good as the congeneric
model.  However, the likelihood ratio test comparing these two models
was significant ($\chi^2_{20} = \Sexpr{round(lrt1[[5]][2],1)}, p=\Sexpr{round(lrt1[[7]][2],2)}$ for the data considered
here).

We can use the proposed ordinal statistics to compare the
tau-equivalent and congeneric models,
fitting the tau-equivalent model and testing whether or
not the factor loadings fluctuate with respect to age group.  In doing
so, we obtain~$\text{WDM}_O=\Sexpr{round(omax1$statistic,1)}, p=\Sexpr{round(omax1$p.value,2)}$
 and~$\text{LM}_O=\Sexpr{round(ol2bb1$statistic,1)}, p=\Sexpr{round(ol2bb1$p.value,2)}$.
Neither statistic is significant at an $\alpha$ of .05,
which supports the conclusions that Froh et al.\ obtained from
alternative fit statistics.  This provides further evidence that the
likelihood ratio statistic is overly sensitive for this particular
analysis.

@@

\subsubsection{GAC} In fitting tau-equivalent and parallel models to the
GAC data, Froh et al.\ obtained mixed results.  The alternative fit
indices did not all agree with one another, and the likelihood ratio test
indicated that the parallel model fit worse than the tau-equivalent
model ($\chi^2_{20} = \Sexpr{round(lrt2[[5]][2],1)}, p<.01$
for the data considered here).  Froh et al.\ ultimately 
concluded that the tau-equivalent model provided a better fit than did
the parallel model.

To apply the ordinal statistics proposed in this paper, we fit the
parallel model and test for instability in the variance parameters
(unique variance and factor variance) w.r.t.\ age.
This results in~$\text{WDM}_O=\Sexpr{round(omax2$statistic,1)}, p<.01$
and~$\text{LM}_O=\Sexpr{round(ol2bb2$statistic,1)}, p<.01$.
Both of these statistics agree with the likelihood ratio test and with
the general conclusion that the tau-equivalent model is better than
the parallel model.  This provides further evidence that the
significant likelihood ratio test is not simply an artifact of the
large sample size.

\subsection{Summary}
The application considered above shows that the ordinal test
statistics can provide useful information in situations where one
might question significant likelihood ratio test statistics.  While
researchers use rules of thumb to obtain decisions from other
alternative fit measures, the proposed statistics are proper tests of
the hypothesis of interest.  They can be used to either supplement or
replace the likelihood ratio test, depending upon the measurement
invariance violations in which the researcher is interested a priori.
We further describe the issue of supplementing vs.\ replacing the
likelihood ratio test in the general discussion.

@@

\begin{enumerate}
  \item To my knowledge, these are the only proposed statistics that
    treat ordinal auxiliary 
    variables as truly ordinal.  LRT treats them as categorical,
    inclusion in the model typically treats tham as continuous.
  \item The first point is important because, if we test w.r.t.\ an
    ordinal auxiliary variable, we are generally looking for ordinal
    violations.  We don't care if a middle level differs from
    all others because we would not usually have an intuitive
    explanation for it.
  \item If we know our sample size is large and have an ordinal
    auxiliary variable, should we even use the LRT in
    the first place?  If the measurement invariance violation is truly
    ordinal, the proposed statistics should be better.  If it is not
    ordinal, the violation will be difficult to explain as anything
    but an anomaly.  The stakes might also play a role here.  In
    high-stakes testing, you might be more interested in treating the
    ordinal variable as categorical to detect any possible
    instability, even those that don't make sense.
  \item The proposed statistics should generally transfer to IRT and
    other psychometric models.
\end{enumerate}
